---
title: "Intro to R"
output: html_document
---
##Setting up our working directory and saving objects vs files

R is a functional programming language, which means that most of what one does is apply functions to objects.

We will begin with a very brief introduction to R objects and how functions work, and then focus on getting data into R, manipulating that data in R, plotting, and analysis.

First, we need to make sure we are working out of the correct directory. You can see teh working directory above the console in RStudio, or use the `getwd()` function to see the current working directory. If you are not currently working in a good location to read and save files, you can use `setwd()` to change the working directory to the location of your choice.

```{r}
setwd("~/Dropbox/Workshops/August_2019_Bioinformatics_Shortcourse_Workshops/intro_to_R/")
```


## Vectors

Let's start by creating one of the simplest R objects, a vector of numbers.
```{r}
v1<-c(1,2,3,4,5)
```

v1 is an *object* which we created by using the ```<-``` operator (less followed by a dash).

v1 now contains the output of the *function* ```c(1,2,3,4,5)``` (c, for **c**ombined, combines elements into a vector

## ditch this
Note that ```=``` is essentially the same as ```<-```, but ```<-``` is preferred by the R community primarily for historical reasons.
```{r}
v1=c(1,2,3,4,5)
```

Let's display the contents of v1:
```{r}
print(v1)
```

We can also just type the name of an object (in this case ```v1```) to display it.

Let's make a few more objects:
```{r}
x<-10
x

y<-11
y

some_letters<-c("a", "b", "c", "d")
```
With this last variable, we had to use ```""``` to specify that we want to create a character vector. Otherwise it thinks we are looking for variables to combine called a,b,c and d.

```{r, eval=FALSE}
some_letters <- c(a,b,c,d)
```


We might want to get a list of all the objects we've created at this point. In R, the function ls() returns a character vector with the names of all the objects in a specified environment (by default, the set of user-defined objects and functions).

```{r}
ls()
```


## Getting Help

R has very good built-in documentation that describes what functions do.

To get help about a particular function, use ? followed by the function name, like so:

?ls

If you are not exactly sure of the function name, you can perform a search:

??mean

We can manipulate objects like so:
```{r}
x+5
x*2
x
```

Note that the value of ```x``` is not modified here. We did not save the output to a new object, so it is printed to the screen.

If we want to update the value of ```x```, we need to use our assignment operator:
```{r}
x<-x+5
x
```

R handles vector math for us automatically:
```{r}
v1*x
v2<-v1*x
v2
```

> **Vector exercises:**
>
>1. Create a new vector,  (called nums), that contains any 5 numbers you like
>
>2. Create a new vector, squares, that contains the elements of nums squared (hint: 2^2 = 2 squared)
>
>3. There are a number of functions that operate on vectors, including length(), max(), min(), and mean(), all of which do exactly what they sound like they do. So for example length(nums) should return 5, because nums is a 5-element vector. Use these functions to get the minimum, maximum, and mean value for the nums vector you created. Also try them on the some_letters vector. What happens?
>
>4. Remember that the function ls() returns a vector of the names of the objects you've created in your current R session. Use ls() and length() to determine the number of objects in your R session.

Note for this last question, you can solve this either by creating a new object (e.g. ```cur_obj```), or by nesting the functions inside each other.

```{r}
#1
nums <- c(33, 22, 41, 54, 91)

#2
squares <- nums^2
squares

#3
length(nums)
min(nums)
max(nums)
mean(nums)

length(some_letters)
min(some_letters)
max(some_letters)
min(some_letters)

#4
length(ls())
cur_obj <- ls()
length(cur_obj)
```

We can also remove objects from our workspace using the function ```rm```.
?rm
```{r}
ls()
rm(x)
ls()
```

We might want to do this if we are working with large object that takes up a lot of memory, and we no longer need that object.


## Object Types...maybe skip

All objects have a type. Object types are a complex topic and we are only going to scratch the surface today. To slightly simplify, all data objects in R are either atomic vectors (contain only a single type of data), or data structures that combine atomic vectors in various ways. We'll walk through a few examples. First, let's consider a couple of the vectors that we've already made.

```{r}
nums <- c(33, 22, 41, 54, 91)
class(nums)
class(some_letters)
```

*Numeric* and *character* data types are two of the most common we'll encounter, and are just what they sound like. Another useful type is *logical* data, as in TRUE/FALSE. We can create a logical vector directly like so:
```{r}
logic1<-c(TRUE, TRUE, FALSE, FALSE)
logic1
class(logic1)
```

Or we can use logical tests.
```{r}
logic2<-v1>2
logic2
```

Note that the logical test here (>2) is applied independently to each element in the vector. We'll come back to this when we talk about data subsets.

A final thing to note about logical vectors: they can be converted to numeric with TRUE=1 and FALSE=0. So a simple way to find how many elements in a logical vector are true is sum(). Using mean() is an easy way to get the proportion of true elements in a vector.

```{r}
sum(logic2)
mean(logic2)
```

> **Object type exercises:** SKIP
>
>1. Use class() to determine the type of vector that the ls() function returns. How about the max function? Does it change depending on the input?
>
>2. Construct a logical vector with the same number of elements as your nums vector, that is TRUE if the corresponding element in the nums vector is less than the mean of the nums vector, and FALSE otherwise. Call this vector big_nums
>
>3. How many elements in your nums vector are greater than the mean of the nums vector?
>

```{r}
#1
class(ls())
class(max(nums))
class(max(some_letters))

#2
big_nums <- nums>mean(nums)

#3
sum(big_nums)
```


## Subsetting and Logical Vectors JUST NUMERICAL INDICES

We can also extract portions of vectors that we've created using ```[]``` following the vector, with the positions we wish to extract. For example, we can extract portions of the nums vector:

```{r}
nums[1]
nums[5]
nums[6]
nums[1:5]
nums[c(1,4,5)]
nums[c(5,5,5,5)]
```

Now it can be useful to specify which positions you would like to include, but often it is more useful to create new objects based on some sort of rule. That is where logical vectors come in, and really how you will mostly be using logical vectors going forward.

For example, in the last exercise, we constructed a logical vector that was the same length as our nums vector that was ```TRUE``` with if the corresponding element in the nums vector is less than the mean of the nums vector, and ```FALSE``` otherwise.

Let's take a look at both of these vectors again.

```{r}
nums
big_nums
```

Let's say we actually want to extract all of the numbers that are bigger than the mean (so all of the ```TRUE```) values. All we have to do is put the logical vector in brackets.

```{r}
nums[big_nums]
```

We don't necessarily have to define this vector ahead of time, but can put the rule in the brackets as well.

```{r}
nums[nums>mean(nums)]
```

> **Vector Subsetting Data exercises:**
>
> 1. Create a logical vector that is true for all values of nums that are larger than the minimum.
>
> 2. Subset nums for all numbers that are larger than the minimum.
>
> 3. Subset nums for all values that are not equal to the maximum of nums.
>
> Hints -- logical tests in R - greater than, less than, equal to, not equal to
>
>  > '> , < , == , !='

```{r}
#1
nums2 <- nums>min(nums)

#2
nums[nums2]

#3
nums[nums != max(nums)]
```


## Data Frames

So far we've been talking about atomic vectors, which only contain a single data type (every element is logical, or character, or numeric). However, data sets will usually have multiple different data types: numeric for continunous data, character for categorical data and sample labels. Depending on how underlying types are combined, we can have four different "higher-level" data types in R:

**Dimensions** | **Homogeneous** | **Heterogeneous**
---------------|-----------------|-------------------
1-D            | atomic vector   | list
2-D            | matrix          | data frame / tibble

We'll focus on data frames and tibbles for today, but lists and matrices can also be very powerful. A data frame is a collection of vectors, which can be (but don't have to be) different types, but all have to have the same length. Let's make a couple of toy data frames.

One way to do this is with the data.frame() function.

```{r}
df1<-data.frame(label=c("rep1", "rep2", "rep3", "rep4"), data=c(23, 34, 15, 19))
df1
class(df1)
```

str() gives lots of information about the data type of the consituent parts of a data frame
```{r}
str(df1)
```
## Skip factors!
Note that label is a *factor* -- this is a special kind of character vector to represent categorical data. By default, when creating a data frame or reading data from a file, R will convert strings (character vectors) to factors. Factors can be useful, but also introduce some pitfalls, which we can come back to if we have time. A good source of information on factors is [here](http://swcarpentry.github.io/r-novice-inflammation/12-supp-factors/).

We can use the function head() to look at part of a a dataframe (or any R object). This can be very useful if you have a very large or long dataframe. You also can control how many lines of the dataframe you view with ```n=#```, although the default is 6.

```{r}
head(df1)
head(df1, n=2)
head(df1, n=10)
```

The summary() function can also be very useful to get a snapshot of the data in your dataframe.

```{r}
summary(df1)
```



###Reading files into R

So far we have been working with small objects we created by hand. A more common way to create dataframes is by reading from a file. There are a few functions to do this in R: read.table() and read.csv() make this easy.

```{r}
bus<-read.table(file="http://software.rc.fas.harvard.edu/ngsdata/workshops/2015_March/mbta_bus.tsv", sep="\t", header=TRUE)
```

Note that R can read data directly from the web -- no need to download the file first.


> **File reading exercise**
>
>1. Explore the bus data frame and see what it contains
>
>2. Try using head() to see the first 20 rows and summary() to see some information about the data.
>
>3. What happens when you use the function View() with the bus object?

```{r}
#1
#bus

#2
head(bus, n=20)

#3
View(bus)
```

head() can be used to increase the number of rows shown when looking at an R object. Let's also look at str()

Again, summary() is a good way to see some basic information about a dataframe

```{r}
summary(bus)
```

The view function can be useful for viewing dataframes in a spread-sheet like format.

```{r, eval=FALSE}
View(bus)
```

### Subsetting and Manipulating Dataframes

We learned about how to subset vectors earlier, now let's look at ways to get subsets of data from two-dimensional data. R uses ```[]``` to index rows and columns, like so:

```{r, eval=FALSE}
bus[1,1]
bus[1,]
bus[15,2]
bus[1:5,]
bus[1:5,4]
bus[c(1,3,5),c(2,3)]
head(bus[,1])
bus[1:6,1]
```

The first number refers to rows, and the second to columns in the case for 2-dimensional objects.

Notice that when you are extracting data from a single column, the data are returned as a vector of the same type as the column. However, if you extract more than one column, the data are returend as a data frame.

```{r, eval=FALSE}
str(bus[1,1])
str(bus[1:5,3])
str(bus[1:3,1:3])
```


Often we want to select a set of rows from a data frame matching some criteria. We can use the subset() function for this. Let's start by just getting the data for key bus routes.

```{r}
key.bus<-subset(bus,type=="Key")
```

Sometimes we might want something more complicated. Let's say we want to get data for all "local" bus routes with at least 1000 daily riders.

```{r}
local.bus<-subset(bus, type=="Local" & ridership > 1000)
```

> **Subsetting Data exercises:**
>
> 1. Subset bus when ridership is greater than 2500.
>
> 2. Subset bus when bus routes are NOT classified as Key.
>
> 3. Subset bus using the conditions of both #1 and #2, save as bus_2500_notKey
>
> Hints -- logical tests in R
>
>  > '> , < , == , !='
>
>  > '& is and -->> z <- x & y TRUE only when both x and y are true'
>
>  > '| is or -->> z <- x | y TRUE if either x or y is true'
>

```{r}
#1
subset(bus,ridership>2500)
#2
subset(bus,type!="Key")
#3
bus_2500_notKey <- subset(bus, ridership>2500 & type != "Key")
```


We can add and manipulate the data frame as well, but we will save that for the next session, and use some packages specifically built for that purpose.

## Writing Data

We've added several variables now, and we might want to write our updated dataset to a file. We do this with the write.table() function, which writes out a data.frame.

```{r}
write.table(bus, file="mbta_bus_data.tsv", quote=F, sep="\t", row.names=F, col.names=T)
```

But where did R put the file on our computer? R does all file operations without a full path in a working directory. RStudio has a preference to set the default working directory, which is typically something like /Users/Allison/R.

To see the current working directory, use:
```{r}
getwd()
```

We can change the working directory with ```setwd()```.

Note that you can also save R data objects directly in an R data file (.Rdat). This can come in handy if you have a number of R objects that took a long time to compute, and you want to work on them later. This can also be useful if you are working with R objects that can't easily be saved as a tab-delimited file.

```{r}
save(bus,bus_2500_notKey,file="bus_objects.Rdat")
rm(bus)
rm(bus_2500_notKey)
```

Later, when you load the file, those objects will automatically be added to your environment.
```{r}
load("bus_objects.Rdat")
```


>**File operators exercises:**
>
>1. Write a copy of bus_2500_notKey to your hard drive, but **be sure to give it a different name than the original!** Otherwise you will overwrite your original data.
>

```{r}
write.table(bus_2500_notKey,file="bus_2500_notKey.tsv",quote=F,sep="\t",row.names=F,col.names=F)
```
